---
title: "Spotify Segmentation"
author: "Hamna Khalid"
date: "2026-02-26"
output: word_document
---



```{r}
library(tidyverse)
```


```{r}
##install.packages(c("tidyverse", "factoextra", "cluster", "janitor"))
library(tidyverse)
library(factoextra)
library(cluster)
library(janitor)
library(GGally)
library(reshape2)
```

Uploading dataset

```{r}
df <- read_csv("dataset.csv")
df
glimpse(df)
dim(df)
names(df)
colSums(is.na(df))
df %>%
  select(where(is.numeric)) %>%
  summary()
```

```{r}

ggplot(df, aes(x = popularity)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Song Popularity",
       x = "Popularity",
       y = "Count")
```


```{r}
ggplot(df, aes(x = energy)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Distribution of Energy",
       x = "Energy",
       y = "Count")

```


```{r}
ggplot(df, aes(x = danceability)) +
  geom_histogram(bins = 30, fill = "orange", color = "black") +
  labs(title = "Distribution of Danceability",
       x = "Danceability",
       y = "Count")

```


```{r}

ggplot(df, aes(x = acousticness)) +
  geom_histogram(bins = 30, fill = "plum", color = "black") +
  labs(title = "Distribution of Acousticness",
       x = "Acousticness",
       y = "Count")
```


```{r}
ggplot(df, aes(x = valence)) +
  geom_histogram(bins = 30, fill = "pink", color = "black") +
  labs(title = "Distribution of Valence",
       x = "Valence",
       y = "Count")

```

```{r}
eda_long <- df %>%
  select(danceability, energy, acousticness, instrumentalness, valence) %>%
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value")

ggplot(eda_long, aes(x = feature, y = value, fill = feature)) +
  geom_boxplot() +
  labs(title = "Boxplots of Key Spotify Audio Features",
       x = "Feature",
       y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r}
corr_data <- df %>%
  select(popularity, danceability, energy, loudness, speechiness,
         acousticness, instrumentalness, liveness, valence, tempo)

corr_matrix <- cor(corr_data, use = "complete.obs")

corr_melt <- melt(corr_matrix)

ggplot(corr_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  labs(title = "Correlation Heatmap of Key Audio Features",
       x = "",
       y = "",
       fill = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


```{r}
ggplot(df, aes(x = acousticness, y = energy)) +
  geom_point(alpha = 0.2, color = "darkblue") +
  labs(title = "Energy vs Acousticness",
       x = "Acousticness",
       y = "Energy")

```


```{r}

df %>%
  count(track_genre, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(track_genre, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Genres in the Dataset",
       x = "Genre",
       y = "Count")

```
Performing PCA

```{r}
pca_data <- df %>%
  select(danceability, energy, loudness, speechiness,
         acousticness, instrumentalness, liveness,
         valence, tempo)

```


```{r}
pca_scaled <- scale(pca_data)

```


```{r}
pca_model <- prcomp(pca_scaled)
summary(pca_model)

```


```{r}
plot(pca_model, type = "l", main = "Scree Plot for PCA")

```


```{r}
library(factoextra)

fviz_pca_biplot(pca_model,
                label = "var",
                repel = TRUE,
                title = "PCA Biplot of Spotify Audio Features")

```


```{r}
summary(pca_model)
```
# Converting  explicit TRUE/FALSE into 0/1 (if it exists)

```{r}
# Converting  explicit TRUE/FALSE into 0/1 
df <- df %>%
  mutate(explicit = ifelse(explicit == TRUE, 1, 0))

features <- df %>%
  select(
    popularity, duration_ms, explicit,
    danceability, energy, key, loudness, mode,
    speechiness, acousticness, instrumentalness,
    liveness, valence, tempo, time_signature
  )

# Handle missing values (if any)
features <- features %>%
  mutate(across(everything(), ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)))


```

Scale the data 
```{r}
X <- scale(features)  # standardize (mean=0, sd=1)
```

choosing K 

```{r}
set.seed(123)
sample_n <- 10000  
idx <- sample(1:nrow(X), sample_n)

X_sample <- X[idx, ]
```

Elbow Method

```{r}
fviz_nbclust(X_sample, kmeans, method = "wss") +
  ggtitle("Elbow Method (WSS)")
```

Sillouhete Method

```{r}
fviz_nbclust(X_sample, kmeans, method = "silhouette") +
  ggtitle("Silhouette Method")

```

choosing K=4 to 8 
running K means 

```{r}
set.seed(123)
k <- 5

km <- kmeans(X, centers = k, nstart = 25)

df_clustered <- df %>%
  mutate(cluster = factor(km$cluster))
```

Visualizing Clusters 
```{r}
fviz_cluster(km, data = X,
             geom = "point",
             ellipse.type = "norm") +
  ggtitle("Spotify Segmentation (K-means clusters)")
```

what does each segment mean?
cluster profiles (mean feature values)


```{r}
cluster_profile <- df_clustered %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    across(c(popularity, danceability, energy, loudness,
             speechiness, acousticness, instrumentalness,
             liveness, valence, tempo, duration_ms),
           mean, na.rm = TRUE),
    .groups = "drop"
  )

cluster_profile


```
Top genres per cluster

```{r}
top_genres <- df_clustered %>%
  count(cluster, track_genre, sort = TRUE) %>%
  group_by(cluster) %>%
  slice_head(n = 10)

top_genres
```

Sample "example tracks" per cluster
```{r}
example_tracks <- df_clustered %>%
  select(cluster, track_name, artists, track_genre, popularity) %>%
  group_by(cluster) %>%
  slice_max(order_by = popularity, n = 5) %>%
  arrange(cluster, desc(popularity))

example_tracks
```





```{r}
cluster_profile
```
Cluster 1 = Acoustic & Soft Listeners
Cluster 2 = High-Intensity / Workout Listeners
Cluster 3 = Upbeat Pop / Party Listeners (largest cluster)
Cluster 4 = Rap / Hip-Hop Focused Listeners
Cluster 5 = Instrumental / Study & Ambient Listeners


```{r}

```




```{r}

```




```{r}

```



```{r}

```





```{r}

```




```{r}

```




```{r}

```

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
